---
title: "Class Reference"
author: "Hanna Zakharenko"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.


## Things I Learned on Jan. 27

The command to set a working directory is setwd(). For example; 
```{r}
setwd("~/Documents/GitHub/data_journalism_2022_spring")
```

## Prelab 2
``` {r}
 #gives the first couple rows of data
head(maryland_ppp)

#gives the values of each field/variable/column (max, min, median, etc.)
summary(maryland_ppp)

#shows the names of all the columns
colnames(maryland_ppp)

#shows column name, data type, and examples of what first couple values look like
glimpse(maryland_ppp)
```


### arrange()
orders the rows of a data frame by the values of selected columns.
``` {r}
arrange(desc(percent_payroll)) #arrange in descending order

arrange((percent_payroll) # default - arrange in ascending order

```

### if_else()
if_else(condition, true, false, missing = NULL)
``` {r}
# if/else example
maryland_with_in_out <- maryland_ppp %>%
  mutate(
    in_out = if_else(
        servicing_lender_state == 'MD', "IN", "OUT"
      )
  )
```


### case_when()
multiple if_else() statements
equivalent of SQL CASE WHEN statemetn
If no cases match, NA is returned
```{r}
# case_when example
maryland_jobs_categories <- maryland_ppp %>%
mutate(
  jobs_retained_category = case_when(
    jobs_retained < 10 ~ 'under_10',
    jobs_retained >= 10 & jobs_retained < 25 ~ '10_to_25',
    jobs_retained >= 50 & jobs_retained < 100 ~ '50_to_100',
    jobs_retained >= 100 & jobs_retained < 250 ~ '100_to_250',
    jobs_retained >= 250 & jobs_retained < 500 ~ '250_to_500',
    jobs_retained == 500 ~ '500'
  )
)
```


## Filters

Filter is a tidyverse verb that uses criteria you supply to return a subset of the original data. Here's how it is structured:

`filter(COLUMN_NAME OPERATOR CRITERIA)`

Where COLUMN_NAME is any column name from your dataframe, OPERATOR is some comparison operator like == or > or < and CRITERIA is something to compare the values in COLUMN_NAME to.




### read.csv vs read_csv
read.csv - Reads a file in table format and creates a data frame from it, with cases corresponding to lines and variables to fields in the file.

read_csv - will make assumptions about the data
```{r}
# using read.csv - base R
maryland_ppp_sample <- read.csv('data/ppp_maryland.csv')
glimpse(maryland_ppp_sample)

# using read_csv from the tidyverse - makes assumptions about the content of the data
maryland_ppp_sample <- read_csv('data/ppp_maryland.csv')
glimpse(maryland_ppp_sample)
```


### lubridate
- can parse functions or turn other columns into date format, but have to match order as much as possible
e.g. dmy(), myd(), dym(), ymd_hms() or parse_date_time()
```{r}
lubridate_maryland_ppp <- maryland_ppp_sample %>% mutate(
    CleanDate = ymd(date_approved)
)
glimpse(lubridate_maryland_ppp)
```

##changing data types
you can change the data type of a column by mutating it and using as.datatype
```{r}
mutate(column_name = as.datatype(column_name))
#e.g.
mutate(id = as.character(id))
```


## extract month and year
```{r}
ppp_maryland_loans %>%
  mutate(month_year_approved = floor_date(date_approved, "month")) %>%
  group_by(month_year_approved) %>%
   summarise(
    count=n()
)%>%
  arrange(month_year_approved)

```



## clean_names() and rename()
clean_names() standardizes column names and rename(new_name=old_name) changes the name of the column
```{r}
# cleaning function
cleaned_arnold_md_loans <- arnold_md_loans %>%
  clean_names()
  rename(id = x1_id)
```


## get_dupes
finds duplicate rows in the database
```{r}
cleaned_arnold_md_loans %>%
  get_dupes()
```


## distinct()
gets rid of duplicate rows
```{r}
# cleaning function
cleaned_arnold_md_loans <- arnold_md_loans %>%
  clean_names() %>%
  rename(id = x1_id) %>%
  mutate(amount = as.numeric(amount)) %>%
  distinct()


# display the cleaned dataset
cleaned_arnold_md_loans

```


## Cleaning data
str_sub() - Cleans up zip codes by turning all 9-digit zip codes into 5-digit zip codes
str_to_title - fixes capitalization of values in column
uses case_when to fix spellings of Arnold to be spelled correctly by looking for strings that are spelled a certain way and replacing them 
```{r}
# cleaning function
cleaned_arnold_md_loans <- arnold_md_loans %>%
  clean_names() %>%
  rename(id = x1_id) %>%
  mutate(amount = as.numeric(amount)) %>%
  distinct() %>%
  mutate(zip = str_sub(zip, start=1L, end=5L)) %>%
  mutate(city = str_to_title(city)) %>%
  mutate(city = case_when(
    city == "Anold" ~ "Arnold",
    str_detect(city,"^Arnl") ~ "Arnold",
    TRUE ~ city
  ))


# display the cleaned dataset
cleaned_arnold_md_loans

```