---
title: "lab_11"
author: "sean mussenden"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. You'll need tidyverse, tidycensus, janitor, sf and tigris, and you'll also need to install your census API key.
```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(sf)
library(tigris)

```

## Load data

**Task** Create a codeblock and load the WV slice of the PPP loan data, which is in the data folder and is called lab_11.rds.
```{r}
wv_ppp <- read_rds("data/lab_11.rds")

```

## Questions

**Q1.** You are interested in doing some reporting on how West Virginia companies in the "Mining, Quarrying, and Oil and Gas Extraction" industry (https://www.naics.com/six-digit-naics/?code=21) used PPP loans during the pandemic (West Virginia is a big coal mining state).  

Then produce a choropleth map that shows West Virginia counties shaded according to the total amount of loans to that industry per 100,000 population.

Use the magma color palette, and shade on a "log" scale. Also produce a table that shows the total amount of loans per 100,000 population for each county, ranked from highest to lowest, along with the total amount of loans and the total number of loans.

Then answer these questions:

* Based on the map, what region of the state appears to have gotten more loans in this industry? What geographic pattern do you notice in the data?
* Based on the table, what county got the highest total loan amount? What was the total amount per 100K population, the total amount of loans (not population adjusted), how many loans were there, and what's the population?

**A1.**Based on the map, the region of the state that appears to have gotten more loans in this industry is clustered around the center of the state, possibly leaning towards the Northwest. It looks like most of the mining, quarrying, and oil and gas extraction seem to be grouped in this area of the state. There is one outlier (Ritchie county) that seems to have a much higher loan amount per 100k than the rest of the state. 

Based on the table, the county that got the highest total loan amount was Harrison county, with 9,919,762.98 dollars in PPP loans. For Harrison county, the total amount of loans per 100k population was 14,607,650.03 dollars, the total amount of loans was $9,919,762.98, there were 59 loans and the population is 67,908.


1. Filter to get all loans with naics codes that start with 21.
```{r}
mining_wv_loans <- wv_ppp %>%
  filter(str_detect(naics_code, "^21"))
```

2. Get number of loans in each county
```{r}
wv_mining_county <- mining_wv_loans %>%
  group_by(project_county_name) %>%
  summarise(total = n(), total_amount = sum(amount))
```

2. Get West Virginia population
```{r}
wv_county_population <- get_acs(geography = "county", 
              variables = c(population = "B01001_001"), 
              state = "WV",
              geometry = TRUE)
```

3. Clean population data to match county loan data
```{r}
wv_county_population <- wv_county_population %>%
  rename(project_county_name = NAME) %>%
  mutate(project_county_name = toupper(str_remove_all(project_county_name,", West Virginia|County"))) %>%
  mutate(project_county_name = str_trim(project_county_name,side="both")) %>%
  rename(population = estimate) %>%
  select(project_county_name, population, geometry)
```

4. Combine the two dataframes
```{r}
wv_mining_county_pop <- wv_county_population %>%
  left_join(wv_mining_county)
```

5. Replace NA with 0 and calculate total number of loans to that industry per 100k people'
```{r}
wv_mining_per_100k <- wv_mining_county_pop %>%
  mutate(total = replace_na(total,0)) %>%
  mutate(loans_per_100k = total/population*100000) %>%
  mutate(loans_amount_per_100k = total_amount/population*100000) %>%
  arrange(desc(loans_amount_per_100k))
```

6. Graph map
```{r}
ggplot() + 
  geom_sf(data=wv_mining_per_100k, aes(fill=loans_amount_per_100k)) + 
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log") 
```
7. Create a table that shows the total amount of loans per 100,000 population for each county, ranked from highest to lowest, along with the total amount of loans and the total number of loans.
```{r}
wv_mining_per_100k %>%
  as_tibble() %>%
  select(project_county_name, total, total_amount, loans_per_100k, loans_amount_per_100k, population)
```

**Q2.** You want to understand WHY the spatial pattern you saw above exists.  So you grab a shapefile with the latitude and longitude coordinates of every mine permit in West Virginia from the state Department of Environmental Protection.

Here's the url: https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json

(It's in a slightly different format than the .shp file in the lab. It's another type of shapefile called "geojson".  It will work just the same, though, despite that difference. You can read it directly into R without manually downloading it, by putting the url in the st_read() function we learned about in this week's prelab.)

This data includes both currently operating mines and old ones that have closed.  You want to examine only mines that have a permit expiration date after Sept. 1, 2021.

Make a map with a point for each one of these mine permits and put a blank map with the outline of West Virginia counties (which you can get from Tigris) behind it.

What spatial pattern do you see? How does it compare to the spatial pattern in the first map you made?

**A2.** This shows the majority of the active mines in West Virginia are in the South portion of the state. This is the oppposite of what I saw in the first question, where on this map it looks like there is an absense of mines in that part of the state.

1. Read in the mine data
```{r}
wv_mines <- st_read("https://tagis.dep.wv.gov/arcgis/rest/services/WVDEP_enterprise/mining_reclamation/MapServer/0/query?where=1%3D1&outFields=*&geometry=&geometryType=esriGeometryEnvelope&inSR=4326&spatialRel=esriSpatialRelIntersects&outSR=4326&f=json")
```

2. Filter to only include mines that have a permit expiration date after Sept. 1, 2021.
```{r}
wv_active_mines <- wv_active_mines %>%
    filter(expiredate > as.Date("2021/09/01"))
```

3. Make a map with tigris and plot the points on it
```{r}
counties <- counties() 

wv_counties <- counties %>%
  filter(STATEFP == "54")
```
4. 
```{r}
ggplot() + 
  geom_sf(data=wv_counties, fill="white") + 
  geom_sf(data=wv_active_mines) +
  theme_minimal() 
```

**Q3.** To confirm the spatial comparison you identified in question 2, let's create a new map that layers the mine location points on top of our PPP choropleth map we created in question 1. What do you notice about the number of mines in the county that has the highest total loan amount per 100K?

**A3.** This map shows that the number of mines in the county that has the highest loan amount per 100k does not match up with the amount of active mines in that area. In Ritchie county, which had the highest total loan amount per 100k, has only one mine in that county with a permit that expires after September 1, 2021 (and it looks like it is on the border between counties).

1. Layer the two maps on top of each other
```{r}
ggplot() + 
  geom_sf(data=wv_mining_per_100k, aes(fill=loans_amount_per_100k)) +
  geom_sf(data=wv_active_mines) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma",trans = "log") 
```

**Q4.** Starting from the original WV ppp dataframe, examine the loans to companies in "Mining, Quarrying, and Oil and Gas Extraction" in the county with the highest total loan amount per 100K (you can use the table you created in question 1 to identify the top county) -- RITCHIE

* What single company accounts for the biggest chunk of the total loan amount?
* What specific industry are they in, based on their 6-digit NAICS code? Look them up on the web. What services do they provide?
* How many jobs did they save, according to the data?
* What's your hypothesis for why a company in their specific industry would need to seek a PPP loan during the pandemic?

**A4.** The single company that accounts for the biggest chunk of the total loan amount is HALL DRILLING,  LLC with $6,866,600. This company has a 6-digit NAICS code of 213111, which means they provide services for Drilling Oil and Gas Wells. According to the data, they retained 387 jobs. A company like this might seek a PPP loan during a pandemic, because they have no way of continuing to employ their people. As described on the NAICS code definition, this industry often revolves around contracting to drill oil and gas wells, and if there is less demand for oil, as there was when no one was driving during the pandemic, then it's less likely that this company would get contracted out at the same rate.

1. Filter to get the loans in Roane county
```{r}
ritchie_mining_loans <- mining_wv_loans %>%
  filter(project_county_name == "RITCHIE")
```

2. Find the highest loan amount
```{r}
roane_mining_loans %>%
  arrange(desc(amount))
```

-30-
