---
title: "lab_06"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* [install Tabula](https://tabula.technology/)
* run existing code as directed (look for **Task**).
* write code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes in the lab_06 folder and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
```

## Get Our PDF

We'll be working with the [Maryland Statistical Handbook](https://planning.maryland.gov/MSDC/Documents/md-statistical-handbook.pdf) from the state Department of Planning. It's not tiny (44MB), and you'll want to download it to a place you'll remember (like your Downloads folder).

**Task** Download the file from the above link and save it to your Downloads folder.

## Setup Tabula

**Task** Start Tabula and load the PDF, then extract a table

Start Tabula, then go to http://127.0.0.1:8080/ in your browser. Click the "Browse" button and find the Statistical Handbook file in your Downloads folder and click "open", and then click the "Import button" in Tabula. This will take a minute or two.

Let's go to page 30 of the PDF, Table 2A, "International Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". This is _net migration_, meaning it shows the total number of people moving to Maryland and its regions/jurisdictions _minus_ the total number of people leaving the same, divided into 12-month periods from July to July. In Tabula, draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV (it should be called `tabula-md-statistical-handbook.csv` by default) to your lab_06/data folder.

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)):

**Task** Write code to read the CSV file into a dataframe called `international_migration`. As in the pre_lab, you may have to skip 1 or more rows and you may want to supply easier to use column names that use underscores.

```{r}
# Write the code to load the CSV file here
international_migration <- read_csv("data/tabula-md-statistical-handbook.csv", skip=2,  col_names=c("State/Jurisdiction", "year2010_2011", "year2011_2012", "year2012_2013", "year2013_2014", "year2014_2015", "year2015_2016", "year2016_2017", "year2017_2018", "year2018_2019", "year2010_2019"))
```

Add a column for the type of migration ("international") and populate it:

**Task** Write the code to add a `migration_type` column to the data and give it the value of "international"

```{r}
# Write the code to add the column here
international_migration <- international_migration %>%
  mutate(migration_type = "international")
```

## Answer questions

**Q1** Which region and county/city _below_ the state level accounted for the largest percentage of international migration for the entire time period? You'll need to add and populate a column representing percent of total using `mutate`.
**A1** The Suburban Washington Region is the region below the state level that accounted for the largest percentage of international migration for the entire time period, accounting for 62.8% of the international migration. Montgomery was the biggest county/city, accounting for 36.8% of the state's international migration.

```{r}
# Write the code to produce your answer here
international_migration<- international_migration %>%
  mutate(percent_of_total = year2010_2019/198996*100)

international_migration %>%
  arrange(desc(percent_of_total))

```

**Q2** Write a sentence or two that describes the data you produced in A1. Try to convey the most important idea.
**A2** The data I produced in A1 shows that the majority of people who are moving to Maryland from international areas are residing near DC. 

**Q3** Which region & jurisdiction had the largest percentage change (either negative or positive) for international migration between July 2018 and July 2017? The formula for percentage change is easy to remember: (New-Old)/Old.
**A3** Calvert saw the largest percentage change for international migration between July 2018 and July 2017, with a 76.9% decrease. The largest positive percentage change for international migration was in Kent, with a 28.6% increase.

```{r}
# Write the code to produce your answer here
international_migration <- international_migration %>%
    mutate(percent_change_17_18 = (year2017_2018-year2016_2017)/year2016_2017*100)


international_migration %>%
  arrange(percent_change_17_18)
```

**Q4** What's your best guess as to why these changes occurred, and in those places in particular?
**A4** Calvert, which saw the biggest decrease, is in the Southern Maryland Region. Kent, which saw the biggest increase, is in the Upper Eastern Shore Region. This change might have occurred because international migration tends to be near DC, so these regions might regularly see a declining international population as it is more common to move near DC, often into communities/areas with other international populations.


## Back to Tabula

**Task** Extract second table from Tabula

Let's go to page 31 of the PDF, Table 2B, "Domestic Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". In Tabula, hit the "Clear All Selections" button and then draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV to your lab_06/data folder. (and rename it to `tabula-md-statistical-handbook-domestic.csv`).

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner:

**Task** Write code to read the CSV file into a dataframe called `domestic_migration`. As in the pre_lab, you may have to skip 1 or more rows and you may want to supply easier to use column names that use underscores. Your work on the international file should serve as a guide.

```{r}
# Write the code to load the CSV file here
domestic_migration <- read_csv("data/tabula-md-statistical-handbook-domestic.csv", skip=2,  col_names=c("State/Jurisdiction", "year2010_2011", "year2011_2012", "year2012_2013", "year2013_2014", "year2014_2015", "year2015_2016", "year2016_2017", "year2017_2018", "year2018_2019", "year2010_2019"))
```

Add a column for the type of migration ("domestic") and populate it:

**Task** Write the code to add a `migration_type` column to the data and populate it with the value of "domestic"

```{r}
# Write the code to add the column here
domestic_migration <- domestic_migration %>%
  mutate(migration_type = "domestic")

```

## Answer questions
**Q5** Which Maryland individual jurisdiction saw the largest net decrease in domestic migration overall?
**A5** The individual jurisdiction that saw the largest net decrease in domestic migration overall was The Suburban Washington Region.

```{r}
# Write the code to produce your answer here
domestic_migration <- domestic_migration %>%
  mutate(percent_of_total = year2010_2019/-159993*100)

domestic_migration %>%
  arrange(desc(percent_of_total))
```

**Q6** How many regions & jurisdictions had net positive migration for July 2017, July 2018 and July 2019 (not just individually but all three together)?
**A6** There are 18 regions and jurisdictions that had a net positive migration for July 2017, July 2018, and July 2019 all together. 

```{r}
# Write the code to produce your answer here
domestic_migration <- domestic_migration %>%
  mutate(change2017_2019 = year2016_2017 + year2017_2018 +year2018_2019)

domestic_migration %>%
  filter(change2017_2019 > 0) %>%
  arrange(desc(change2017_2019))
```

**Q7** How would you describe this data? Is there a county or region that stands out, and why?
**A7** I would describe this data as a way to see which areas are growing consistently and which ones are not. I think the domestic data would be very interesting to compare to the international data. It is interesting that Frederick county saw double the amount of domestic migration than the Southern Maryland Region did as a whole. This is interesting because Frederick County is in The Suburban Washington Region, which saw the largest net decrease in domestic migration overall. Calvert, which saw the largest percentage decrease in international migration between 2017 and 2018, is in the Southern Maryland Region, which saw the second largest domestic positive migration between 2017 and 2019. It is also interesting that Kent came up earlier as having the largest percentage increase in international migration between 2017 and 2018, and Kent has the lowest net positive increase in domestic population between 2017 and 2019. 
