---
title: "lab_05"
author: "Sean Mussenden"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and janitor.

**Task** load these two packages.

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)
library(janitor)
```

## Load Data

You'll need to load three data sets for this:

* The West Virginia slice of the PPP loan data (lab_05.rds).
* A "lookup table" that allows you to translate NAICS (industry) numeric codes to industry titles (naics_codes.csv).
* A table of West Virginia population by county (American Community Survey, 2019 5-year averages) (wv_population_county.csv).

All three data sets are in the data folder.  Write code to load the three in the codeblock below.

**Task** Create a codeblock below this task, then read the data sets in in and assign them to appropriate variable names. There's a fourth data set you'll use in this lab, too, of selected loans in ZIP Code 25401. But there's no need to load it now.
```{r}
wv_loans <- read_rds("data/lab_05.rds")

naics_codes <- read_csv("data/naics_codes.csv")

wv_population <- read_csv("data/wv_population_county.csv")

```
## Answer questions

**Q1.** In the data folder, there is a csv called zip_25401_loan_sample.csv.  It contains a sample of loans from West Virginia ZIP Code 25401.

As we read earlier this semester, [multiple loan applications coming from multiple businesses at the same residential street address](https://www.nytimes.com/2021/08/17/business/ppp-fraud-covid.html) might point to fraud. Or it could alert us to companies that used [multiple corporate entities to get more money than envisioned](https://www.propublica.org/article/different-names-same-address-how-big-businesses-got-government-loans-meant-for-small-businesses) by the original law that authorized the program.   

You are going to examine this data to see if we can find a lot of loans coming from the same address.  Here's the problem: the street address field is pretty messy.  The same address appears with minor variations --  "1003 Sushruta Dr" vs "1003 SUSHRUTA DR" -- that will prevent proper grouping.

First, upload the data into Open Refine and standardize/clean the address field - make sure to make a copy of that column to work on and to try all of the options for clustering. If you've done it properly, you should have 65 discrete addresses.

Then export the data from Open Refine, and move it to the proper folder.

Next, load it in the codeblock below, assigning it to an appropriate variable name.

Then answer these questions:
* What is the street address in this data that has the most loans?
* How many loans are there at that street address?
* What are the names of the businesses at that address?

**A1.**
The street address in this data that has the most loans is 126 E Burke St. There are 6 loans at this street address. The names of the businesses at that address are DREW HOLDINGS LLC, BRIX27 LLC, ABRAHAM ASHTON, HUB CO-OP LLC (x2), and RONIN PROPERTIES LLC. 
```{r}
zip_25401 <- read_csv("data/wv_zip_25401.csv")

zip_25401 %>% 
  group_by(address) %>%
  summarise(
    count=n()
  ) %>%
  arrange(desc(count))
```
```{r}
zip_25401 %>%
  filter(str_detect(address, 'Burke'))
```
**Q2.** Do some web research on the businesses that answered question 1.  

Google the street address.  Google the business names and search for their corporate records on [Open Corporates](https://opencorporates.com/). Be sure to find the website of the name of the company that appears twice in the list of businesses and develop an understanding of what it does.

Based on your research, does it seem suspicious that this collection of businesses all got loans using the same address? Why or why not. Use specific facts identified in your research to support your position.

**A2.** Based on these collection of businesses, it seems suspicious that they all got loans using the same address, because it seems to be filed by the same person, Abraham Ashton and Robert Johnson (https://opencorporates.com/companies/us_wv/338359) . When looking at the location on Google Maps, it looks like the only business housed at that address is Hub Co-op, which is a bike company. It is also interesting to see that this company got two loans for the same business name at the same address, with the only difference being a comma between the name and LLC. However, they could be the same people who own multiple businesses (a.k.a. entreprenuers) at the same address, and therefore filed loans for all of their businesses.

**Q3.** Start by using the West Virginia slice of the PPP loan data that you loaded at the start of the lab to create a subset of PPP loans in West Virginia's second largest county (which you can find in wv_population_county). And then use that table you created to answer the following questions:

* Which city in that county had the highest number of loans?
* In that city, which industry title had more loans than any other industry title?

Requirement: you MUST export a dataframe of PPP loans from R Studio at some point in the process (not necessarily at the beginning!), load it into Open Refine, clean the city column, export it from Open Refine, and reimport into R Studio. To export data from R Studio, you will use the write_csv() function, which works like this if we're using the wv_population_county dataframe:

```{r}
# write_csv accepts 2 arguments: the dataframe and the path to the file you write out
write_csv(wv_population_county, "data\wv_population_county.csv")
```

Guidance: there are a lot of steps you'll need to take to answer this question. You may find it helpful to write out in English what you plan to do step-by-step before you start writing code.   

**A3.** The second largest county in West Virginia, Berkeley, had the highest number of loans in Martinsburg, with a total of 1,166 loans. In Martinsburg, the industry title that had more loans than any other industry title was Full-Service Restaurants. 

Step 1: Find the second biggest county in WV by population. This is: Berkeley
```{r}
wv_population %>%
  arrange(desc(population_2019))
  
```
Step 2: Create a subset of PPP loans in West Virginia's second largest county (Berkeley).
```{r}
berkeley_loans <- wv_loans %>%
  filter(project_county_name == "BERKELEY")
```

Step 3: Write out this dataframe as a csv to clean in Open Refine
```{r}
write_csv(berkeley_loans, "data/berkeley_loans.csv")
```

Step 4: Re-load cleaned city data into R
```{r}
clean_berkeley_loans <- read_csv("data/clean_berkeley_loans_csv.csv")
```

Step 5: Answer the questions using cleaned data.
```{r}
clean_berkeley_loans %>%
  group_by(city) %>%
    summarise(
    count=n()
  ) %>%
  arrange(desc(count))
```
```{r}
berkeley_pp_with_naics <- clean_berkeley_loans %>% left_join(naics_codes, by="naics_code")
```

```{r}
berkeley_pp_with_naics %>%
  filter(city == "Martinsburg") %>%
    group_by(title) %>%
    summarise(
    count=n()
  ) %>%
  arrange(desc(count))
```
**Q4.** What are your two best hypotheses, which you'd need to confirm with reporting and further analysis, that explain why that industry is at the top of the list?

**A4.** It is not surprising that restaurants are the top industry in Martinsburg to receive the most loans. When taking a look at Google Maps for this area, it is visible that this town has a lot of hotels and restaurants [https://goo.gl/maps/Xc1ubGGi5wdnSkuT6], and that it is a highly packed area with businesses. It is in an area that seems to be targeted for tourism [https://www.travelwv.com/]. Another hypothesis is that it is also known that full-service restaurants were struggling during COVID because they were unable to open to dine-in or full-capacity dine-in for a long time [https://www.restaurantdive.com/news/paycheck-protection-program-2021-business-expense-deductions/592891/]. Therefore, Martinsburg might have had a lot of restaurants file for PPP checks because they were struggling during the pandemic, and located in a larger business-centered/touristy area.

**Q5.** Start with a dataframe of loans to all businesses in the city and industry that answered question 3. Answer the following questions:
* What is the name of the business that got the highest approved loan amount?
* How much was it for?
* When was it approved?
* How many jobs does the data say were retained?
* Is there a difference between the business' name in the PPP data and the name its customers know it by? If so, what is that name?
* How many locations does this business have?
* Did one of its locations close during the pandemic, either before or after it got the loan?

Hint: you will not find the answers to the last three questions in the data.  You could call them directly to get that information, but I don't want you to do that for this assignment.  Instead, do some web research. I would start by Googling the company name from the data and looking at the page that comes up for the business from at http://apps.sos.wv.gov/. I would use information I found on that page and use info about the company from Google, the [Wayback machine](https://archive.org/web/) (which lets you look at older versions of a company's website), Yelp, and Facebook.

**A5.** The name of the business that got the highest approved loan amount was COTTLE CHRISTI L LLC. It was for $280,434.00. It was approved on	February 17, 2021. It says that 94 jobs were retained. There is a difference between the business' name in the PPP loan data and the name its customers know it by. Their customers know it as Kitzie's Cafe [https://apps.sos.wv.gov/business/corporations/organization.aspx?org=338507]. According to their website, they have two locations, one in Spring Mills and one in Martinsburg [http://www.kitziesrestaurant.com/]. They originally had 3 locations, with one in Inwood as well, but it closed before it got the loan [https://web.archive.org/web/20171116112157/http://www.kitziesrestaurant.com/].After looking through their website archive, Facebook, and Yelp, it looks like this location closed in 2019.
```{r}
martinsburg_restaurants <- berkeley_pp_with_naics %>%
  filter(city == "Martinsburg" & title == "Full-Service Restaurants")
```

```{r}
martinsburg_restaurants %>%
    group_by(name) %>%
    summarise(
    amount=sum(amount)
  ) %>%
  arrange(desc(amount))
```
```{r}
martinsburg_restaurants %>%
  filter(name == "COTTLE CHRISTI L LLC")
```


